---
title: "stringr and dplyr"
author: "Lusine Zilfimian"
date: |
     `r format(Sys.Date(), '%B %d (%A),  %Y')`
output: 
    beamer_presentation:
      theme: "AnnArbor"
      colortheme: "beaver"
      fonttheme: "structurebold"
      incremental: True
      fig_width: 5.5
      fig_height: 3.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
pacman::p_load(ggplot2, ggthemes, FNN, dplyr, stringr, RColorBrewer, wordcloud)
```

# Contents

 * Intro Glossary
 * String Manipulation
 * Stringr vs Base
 * RegEx
 
 ---
 
# Bibliography 

 * See on GitHub: Handling and Processing Strings 
 * Character Strings - Chapter 2
 * String Manipulation + stringr - Chapter 3, 4
 * RegEx - Chapter 5
 
 ---
 
# Intro Glossary

 * String manipulation - series of functions used to extract information from text variables.
 
 * **Regex** is set of commands used to match a family of text (alphanumeric, digits, words) to detect string sequences in a large text data. 
 
 * *Regular expressions* to do more complicated tasks such as extract email IDs or date from a set of text.

 * String manipulation functions are not customized however we can customize regular expressions in any way we want.
 
#  String Manipulation

 * We can create strings with either single quotes or double quotes. 
 
 * Unlike other languages, there is no difference in behavior. 
 
 * To include a literal single or double quote in a string you can use `\` to escape it.
 
```{r}
(x <- "This is my first character!") # double quotes 

class(x) == typeof(x)
(x <- 'This is my second string') # single quotes

(x <- "Why do we need the 'python'?")

(x <- 'Why do we need the "python"?')
```

 ---

## Working with strings

 * If you try to include single quotes within single quotes, or double quotes within double quotes, it will not work

```{r error=TRUE}
# Something went wrong
"Why call langauge "python" ? "
'Why call langauge 'python' ? '
```

 *  `\` escaping character
 
```{r}
(x <- 'Why do we need the \'python\'?')
is.character(x)

"Why do we need the \"python\"?"
```

 ---

```{r}
length(character(0))
length("")
length(character(1))
a = character(3)
a[2] <- "Something"
length(a)
```
 ---
 
* Mixing different type of objects

```{r}
# vector with numbers and characters
c(1:5, pi, "text")
#  matrix with numbers and characters
rbind(1:5, letters[1:5])
# dataframe withnumbers andcharacters
df1 =data.frame(numbers= 1:5,letters = letters[1:5])
str(df1)
# listwith elements of different mode
list(1:5,letters[1:5], rnorm(5))
```
 
 ---

* Paste() and its arguments 
* Takes or more R objects, conver them to character and then concatenates them to character (one or more)
* Element wise concatenation of two vectors 
 
```{r}
x <- 1:26
(a <- LETTERS[x[1:13]])
(b <- letters[x[14:26]])
```

 ---


```{r}
(char1 <- paste(x, c(a, b)))
length(char1)
```

 * sep = The element which separates every term. 
 * collapse = The element which separates every result.


```{r}
paste(x, c(a, b), sep = " - is - ") # optional - Define your own separator
(char2 <- paste(x, c(a, b), sep = " - is - ", collapse = " , ")) # optional -  if you want all the terms to be collapsed into a single string
length(char1); length(char2)
```

\tiny difference between `paste` and `paste0` function - `paste0` does not have the argument sep, and it concatenates the stings objects without spaces. 
 
 
```{r}
paste("Life of", pi)
paste("Life of", pi, sep = "-")
paste("Life of", pi, sep = "-", collapse = ",")

paste(c("Life of", "Other"), c(pi, 9), sep = "-", collapse = " ,")
paste("X", 1:5, sep= ".")
paste("thevalue of 'e' is", exp(1), NA) # problem
?paste0
cat("Life of", pi)
x <- paste("Life of", pi)
```

 * C-stylestringformattingwithsprintf()
 
```{r}
# decimal notationwith 3 decimal digits
sprintf("%.3f",pi)
# with sign(positive)
sprintf("%+f",pi)
?sprintf
#+ stings
```

 ---

# Basic Sting manipulations (Base)

 * nchar() counts the number of characters (including whitespaces) 
 * for single character
 
 
```{r}
nchar("Here are 16 chars:)")
char1
char2
length(char2)
nchar(char1)
```

 ---
 
 * We can convert all characters of our string to lower case with the `tolower` command
 
 * and to upper case with `toupper`

```{r}
toupper("Here are 16 chars:)")
tolower(toupper("Here are 16 chars:)"))

chartr(old = "is", new = "is not", char2) # the size should be equal
chartr(old = "is", new = "no", char2) 


txt <- "Café Münchner Kindl"
clean <- chartr("áéíóúñüäö", "aeiounuao", tolower(txt))
clean
```

 ---

# Substring + replacement based on the positions

 *  substr() and substring()
 *  These function can be used to overwrite or replace a part of character string.

```{r}
x <- "Lusine Zilfimian"
substr(x, start = 8, stop = 7 + 5) 
substring(x, first = 8) #  last = 1000000L
substr(x, start = 14, stop = 14) <- "y"  
x
```

 ---
 
# Replacement and detection

```{r}
sub("Lusine Zilfimian", pattern = "i", replacement = "y") # find and replace
gsub("Lusine Zilfimian", pattern = "i", replacement = "y")

grep(pattern = "Lusine", x =  c("Lusine Zilfimian", "David", "Mar")) #detect
grepl(pattern = "Lusine", x = c("Lusine Zilfimian", "David", "Mar"))
```
```{r}
# Vectors
x <- c("ID1", "ID2", "ID2", "ID3", "ID5")
y <- c("ID2", "ID3", "ID4", "ID4", "ID6")

# Set union vs intersection vs difference
union(x, y)         # unique in x ∪ y
intersect(x, y)     # unique in x ∩ y
setdiff(x, y)       # unique in x \ y
setequal(x, y)      # TRUE (same unique elements)
identical(x, y)     # FALSE (order/length differ)
is.element("ID1",y)
```
 ---

## Why is `base` not enough?
 
 * The object `NULL` have zero length, yet when included inside `paste()` they are treated as an empty string:

```{r}
#c("one", "two", "three", NA, "five") - previous version
paste("This", "is", NULL, NULL, NULL, NULL, "value")
length(NULL)
length(character(0))

stringr::str_c("This", "is", NULL, NULL, NULL, NULL,  "value", sep = " ")
```

 * `str_c()` – zero argument are silently removed
 
 ---

# Stingr 

 * All functions start with str_
 * All functions take a vector of strings as the first argument

 * `str_sub` works the same way as `substr()` but allows for negative subsetting as well

```{r}
library(stringr)
str_sub("Joe Satriani", start = 5)
str_sub("Joe Satriani", start = 5, end = 5+4)
str_sub("Joe Satriani", start = -8, end = -4)
```

 ---

 * `str_detect()` - For detecting whether a pattern is present (or absent) in a string vector. You get a TRUE if a match is detected in a string, FALSE otherwise
 
 * `str_subset()` - Keep strings matching a pattern
 
 * `str_count()` - Count the number of matches in a string
 
 * `str_split()` - Similar to `strsplit()` to separate a character vector into a number of pieces
 
 * `str_replace()` - For replacing the first occurrence of a matched pattern in a string
 
 * `str_exctact()` - For extracting a string containing a pattern 

\tiny detect - bolean, extract – only pattern, subset – all the text.

 ---
 
# Examples
 
```{r}
(geners <- c("Action, Adventure, Comedy",
             "Comedy",
             "Comedy, Drama, Drama, Romance", 
              "Crime, Drama, History", "Action 2"))

str_detect(string = geners,  pattern = "Drama")
str_subset(string = geners,  pattern = "Action")

str <-c("123abd", "lz@yahoo.com", "Name567cd", "abc5.00", "lusinezilfimian@yahoo.com")
str_subset(str, pattern="@yahoo\\.com")
str_count(string = geners,  pattern = "Drama")
```

 ---
 
# Examples
 

```{r}
geners
str_replace(string = geners, pattern = ",", replacement = " &")
sub(x = geners, pattern = ",", replacement = " &") # from base 
gsub(x = geners, pattern = ",", replacement = " &") # from base 
str_replace_all(string = geners, pattern = ",", replacement = " &")
```

 ---
 
# Examples
 
 
```{r}
str_extract(string = geners, pattern = "Drama")
str_extract_all(string = geners, pattern = "Drama", simplify = T)
```

 ---
 
# Let's consider the example

```{r}
movies <- read.csv('movies3.csv', stringsAsFactors = F)
View(movies)
```

## How many romcom movies are there? 

```{r}
movies$Comedy <- str_detect(string = movies$Genre, pattern = "Comedy") 
movies$rom <- str_detect(string = movies$Genre, pattern = "Romance") 

table(movies$Comedy,movies$rom)
```

 ---

# Split string into pieces

```{r}
geners
(gen_m <- str_split(geners, ",", simplify = T))
trimws("  Lus    ine   ") # trim whitespaces
str_trim("  Lus    ine   ")

# somefactor
some_factor= factor(c(1,1, 1,2, 2, 2),labels =c("good", "bad"))
some_factor
nchar(some_factor)
str_length(some_factor)
```

```{r}
# quote(by DouglasAdams)
some_quote = c(
"I may not havegone",
"where I intendedto go,",
"but I think Ihave endedup",
"where I needed tobe")

some_quote = paste(some_quote,collapse =" ")

cat(str_wrap(some_quote,width =30))
cat(str_wrap(some_quote,width =30, indent= 2))

```

```{r}
# somesentence
change= c("Be the change", "you want to be")
# extract first word
word(change,1)
# extract second word
word(change,2)
```

 ---
 
# Regular Expressions

  * Sequance of characters that define a search pattern
  * For instance, one could have the following pattern: 2 digits, 2 letters and 4 digits
  * ?regex - to see more
  
## Form of Regular Expressions
  
  * 1. Literal characters which are characters that match themselves
  * 2. Metacharacters - any character that is not a literal character   . ^ $ + * ? \ | (), {}, [], -
  * 3. Sequences \d, \w, \W, \D 
  * 4. Character classes [A-Z], [0-9]
  * 5. Quantifiers {2}
  * 6. POSIX character classes
  
  * If you want to match the metacharacter you need to use \\ (escapes) before the metacharacter

 ---
 
# Literal and Metacharacters 

 * Literal character - the simplest form of regular expressions are those that match a single charact
 * Metacharacters - special characters that have a reserved status:
 
```{r}
str <-c("Lusine1995^", "Say$1995-02(", "David.", "I*02", "I&95", "1995.27")
str_subset(string = str,  pattern = "Lusine") # literal 
```


```{r}
str_subset(string = str,  pattern = "Lusine|David|Say")
```

```{r}
str_subset(string = "money$",  pattern = "money\\$")
str_subset(string = "This is money23",  pattern = "money$")
```

# Metacharacters 

```{r}
str
str_subset(string = str,  pattern = ".") # meta 
str_subset(string = "\n",  pattern = ".") # meta
str_subset(string = " ",  pattern = ".") # meta
str_subset(string = "",  pattern = ".") # meta
str_subset(string = str,  pattern = "\\.") # literal
```

 
# Metacharacters 
 
```{r}
str
str_subset(string = str,  pattern = "^S") # meta
str_subset(string = str,  pattern = "\\^") # literal
str_subset(string = str,  pattern = "\\.") # literal

# str_subset(string = c("'\a'", " '\\' "),  pattern = "'\'") # literal
```


# Sequences

```{r}
str_subset(string = c("123", "4L", "M"),  pattern = "\\d") # [0-9]
str_subset(string = c("123", "4L", "M"),  pattern = "\\D") # [^0-9]

str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s") 
str_subset(string = c(" ", "l ", "l"),  pattern = "\\S") 
```

# Character classes

```{r}
str
str_subset(string = str,  pattern = "[0-9]")
str_subset(string = str,  pattern = "[f-z]")
str_subset(string = str,  pattern = "[a-zA-Z]")
str_subset(string = str,  pattern = "[aeuio]")
str_subset(string = c("AI", "ai"),  pattern = "[AEIOU]")
```

# Special meaning of ^

 * Not in specified character 
 
```{r}
str_subset(string = c("ou", "lk"),  pattern = "[^aeuio]")
str_subset(string = c("ou", "19", "-"),  pattern = "[a-zA-Z0-9]")
str_subset(string = c("ou", "19", "-"),  pattern = "[a-zA-Z]")
str_subset(string = c("ou", "19", "-"),  pattern = "[^0-9]")
str_subset(string = str,  pattern = "[^1-9]")
str_subset(string = c("T", "adfjh", "tdkfj", "TtT"),  pattern = "[tT]")
```

# Quantifiers

```{r}
str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s+") 
str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s*") 
str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s?") 

str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s{3}")
str_subset(string = c(" ", "l   ", "l"),  pattern = "\\s{2,}")
```

# POSIX character classes

```{r}
str_subset(string = str,  pattern = "[:alnum:]") # [a-zA-Z0-9]
str_subset(string = str,  pattern = "[:alpha:]") # [a-zA-Z]
str_subset(string = str,  pattern = "[:digit:]") # [0-9]
str_subset(string = str,  pattern = "[:lower:]") # [a-z]
```

# POSIX character classes

```{r}
str_subset(string = str,  pattern = "[:upper:]") # [A-Z]
str_subset(string = str,  pattern = "[:space:]") # [\f\n\r\t\v]
str_subset(string = str,  pattern = "[:punct:]")
```


```{r}
str_subset(string = c("grey", "gray"), pattern = "gr(e|a)y")
```

 ---
 
# How many wins and nominations are there?

```{r}
movies$awards_num <- str_replace_all(movies$Awards, pattern="[^0-9]",
  replacement = " ")

movies$awards_num <- str_replace_all(movies$awards_num, pattern="\\s+",
  replacement = " ")
movies$Awards[1:10]
movies$awards_num[1:10]
```

 ---
 
# How many wins and nominations are there?

```{r}
movies$awards_num <- trimws(movies$awards_num)
movies$awards_num[1:20]
```

```{r}
x1 <- str_split(movies$awards_num, pattern = "\\s",simplify = T)
x1 <- apply(x1, 2, as.numeric)
x1 <- rowSums(x1, na.rm=T)
x1[1:20]
```

 ---
 
## Visualization 

```{r}
movies$awards_num <- x1
ggplot(movies, aes(x = awards_num)) + geom_histogram() + 
  xlim(c(0,100)) + labs(title = "Awards histogram (wins and nominations)",
    x = "Awards", y = "Count")
```

 ---

#  Wins only 

 * Possible strings are ...
 
```{r}
m <- movies$Awards[1:20]
m1 <- str_extract_all(movies$Awards, pattern="[0-9]+\\s+win", simplify = T)
m1 <- str_remove_all(m1, pattern="[a-zA-Z]")
m1 <- as.numeric(m1)
movies$Wins <- m1
head(m1)
```


 ---
 
## Visualization 

```{r}
ggplot(movies, aes(x = Wins)) + geom_histogram() +
  labs(title = "Wins histogram", x = "Wins", y = "Count")+
  xlim(c(0,100))
```

# Dplyr - Grammar of data manipulation

```{r}
library(dplyr)
library(nycflights13)  # flights dataset: columns include year, month, carrier, etc.
library(tm)
```




```{r setup, message=FALSE, warning=FALSE}
# Load packages
library(dplyr)
# Use mtcars with rownames as a column
df <- as_tibble(mtcars, rownames = "model")
```

## Select / Rename / Relocate

```{r}
# Select specific columns and helpers
s_select <- df |>
  select(model, mpg, cyl, contains("p"))  # contains() picks columns with 'p'

# Rename columns
s_rename <- df |>
  rename(horsepower = hp, gears = gear)
```

## Filter / Arrange

```{r}
# Row filtering with logical conditions
f_basic <- df |>
  filter(mpg > 25, cyl == 4)

# if_any / if_all predicates across multiple columns
f_if_any <- df |>
  filter(if_any(c(mpg, hp), ~ . > 200))  # any of mpg or hp > 200

f_if_all <- df |>
  filter(if_all(c(cyl, gear), ~ . >= 4))  # all of cyl and gear >= 4

# Range filtering
f_between <- df |>
  filter(between(mpg, 20, 25))

# Sorting
a_sort <- df |>
  arrange(cyl, desc(mpg))
```

## Distinct / Deduplication

```{r}
# Unique combinations
d_pairs <- df |>
  distinct(cyl, gear)

# Keep the first row per unique combination (retain all columns)
d_keep_all <- df |>
  distinct(cyl, gear, .keep_all = TRUE)
```

## Mutate / Transformations

```{r}
# Create new columns
m_basic <- df |>
  mutate(pwr = hp * drat, ratio = mpg / wt)

# Conditional label
m_case <- df |>
  mutate(class = case_when(
    hp >= 200 ~ "high",
    hp >= 150 ~ "mid",
    TRUE ~ "low"
  ))


# lead/lag and differences
m_leadlag <- df |>
  mutate(mpg_lead = lead(mpg),
         mpg_lag  = lag(mpg),
         mpg_diff = mpg - lag(mpg))

# Cumulative sum by group
m_cumsum_by_cyl <- df |>
  mutate(cum_hp = cumsum(hp), .by = cyl)
```


## Summarise 

```{r}
# Global summary
s_global <- df |>
  summarise(mean_mpg = mean(mpg), sd_mpg = sd(mpg))
```

```{r}
# Grouped summary (using .by for brevity)
s_by_cyl <- df |>
  summarise(n = n(),
            models = n_distinct(model),
            mean_mpg = mean(mpg),
            .by = cyl)

# Multiple quantiles by group
s_quant <- df |>
  summarise(q = quantile(mpg, c(.25, .5, .75)),
            p = c("q25", "q50", "q75"),
            .by = cyl)
```

## Slice Helpers

```{r}
# Top/bottom/sample rows
sl_head <- df |> slice_head(n = 5)
sl_tail <- df |> slice_tail(n = 5)
sl_samp <- df |> slice_sample(n = 5)

# Min/Max within groups
sl_min_by <- df |> group_by(cyl) |> slice_min(mpg, n = 2) |> ungroup()
```

## Column Select Helpers

```{r}
# Type-based select
sel_numeric <- df |> select(where(is.numeric))

# Pattern-based select
sel_patterns <- df |> select(starts_with("c"), ends_with("p"), contains("ar"))

# Reorder with everything()
sel_everything <- df |> select(-drat, everything())
```

## Joins

```{r}
# Key table for joins
key <- tibble(cyl = c(4, 6, 8),
              cyl_label = c("four", "six", "eight"))

j_left  <- df |> left_join(key, by = "cyl")
j_inner <- df |> inner_join(key, by = "cyl")

df1 <- data.frame(name = c("John", "Paul", "Mick"), band = c("Beatles", "Beatles", "Stones"))
df2 <- data.frame(name = c("John", "Paul"), instrument = c("Guitar", "Bass"))

# Semi join: Keep rows in df1 where 'name' matches in df2
semi_join(df1, df2, by = "name")

df1 <- data.frame(
id = c(1, 2, 3, 4),
value = c(10, 20, 30, 40)
)
df2 <- data.frame(
id = c(2, 4),
value = c(20, 40)
)
anti_join(df1, df2, by = "id")
```

## Bind Rows / Bind Cols

```{r}
# Create two compatible tables
df_A <- df |> select(model, mpg, cyl) |> mutate(src = "A")
df_B <- df |> select(model, mpg, cyl) |> mutate(src = "B") |> slice_sample(n = 10)

# Stack rows
b_rows <- bind_rows(df_A, df_B)

# Bind columns (must align by row order)
b_cols <- bind_cols(df |> select(model, mpg), df |> select(hp, wt))
```


## Group Manipulation 

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(N=n(), Mean = mean(hp))
```

```{r}
mtcars |>
  group_by(cyl) |>
  summarise(N=n(), Mean = mean(hp))
```

