---
title: "Lesson 5"
author: "Lusine Zilfimian"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(dplyr, ggplot2, chron, stringr)
```

# Contents

 * Basic Visualization (base R graphics or traditional R graphics)
 * ggplot

# Basic Plotting

 *  Overview of the `plot()` function 
 * `ggplot2` -  a powerful library for visualizing data in R

## Base R graphics or traditional R graphics

* The R function `plot()` takes in two vectors—one vector of x locations and one vector of y locations—and opens a graphics device where it displays the result.

```{r}
year <- c(2014, 2015, 2016, 2017, 2018, 2019, 2020)
energy_to_learn <- c(0.6, 0.5, 0.1, 0.7, 0.99, 0.5, 0.8) 

plot(x = year, y = energy_to_learn)
```


 * Coordinates in the form of a *matrix*, with x values in the first column and the y values in the second columns 


```{r}
(mat <- cbind(year, energy_to_learn))
plot(mat)
```

```{r}
?plot
```

### Graphical parameters (type, main, xlab, ylab, col, pch, cex, lty, lwd, xlim, ylim)

 * Specify a single character-valued option for the argument `type`

```{r}
plot(mat, type = "l") # lines only
plot(mat, type = "p") # point only 
plot(mat, type = "b") # both points and lines
plot(mat, type = "o") # overploting points

plot(mat, type = "h")
plot(mat, type = "s")
```

 * Title and Axis Labels

```{r}
plot(mat) # by default

plot(mat, type = "p", 
     main = "The title \n with new line", 
     xlab = "This is x", ylab = "This is y")
```


 *  Colors to make data much clearer by distinguishing factor levels or emphasizing important numeric limits, etc.

 * Character string or integer value 

```{r}
colors()
plot(mat, type = "b", main = "The title", xlab = "This is x", ylab = "This is y", 
     col = "black")

plot(mat, type = "b", main = "The title", xlab = "This is x", ylab = "This is y", 
     col = "blue4")

colors()[1:100]

plot(mat, type = "b", main = "The title", xlab = "This is x", ylab = "This is y", 
     col = 1) # black
?plot
plot(mat, type = "p", main = "The title", xlab = "This is x", ylab = "This is y",
     col = c(1,2,3,4,5,6,7))

plot(mat, type = "p", main = "The title", xlab = "This is x", ylab = "This is y", 
     col = c(1,2,3))
```

```{r}
plot(mat, type = "p", main = "The title", xlab = "This is x", ylab = "This is y", 
     col = 2)
```

* `pch` - point character (1-25 or single character)
* `cex` - character expansion (default is 1)

```{r}
plot(cbind(1:25, 1:25), type = "p", pch = 1:25, col = colors()[25:50], cex = 1) # 1-25

plot(cbind(1:25, 1:25), type = "p", pch = "0", col = colors()[25:50], cex = 1:2)
```

 * Line and Point Appearances

* `lty` - line type (1-6 or solid, dotted, dashed, etc.)
* `ltw`- line width


```{r}
plot(mat, type ="b",
     main = "Title", xlab = "", ylab = "", 
     lty = 3, cex = 1.5, lwd = 2) 

plot(mat, type ="b",
     main = "Title", xlab = "", ylab = "", 
     lty = "dashed", cex = 0.5, lwd = 1) 
```

 * Plotting Region Limits

* By default R sets the range of each axis by using the range of the supplied x and y values (plus a small constant to pad a little area around the outermost points).

```{r}
plot(mat, type = "l", xlim = c(1,50)) # c(lower,upper)
plot(mat, type = "l", xlim = c(2014,2021)) 
plot(mat, type = "l", ylim = c(0,2)) 
```

 * Adding Points, Lines, and Text to an Existing Plot 
 * `points()`, `abline()`, `segments()`, `legend()`, `text()`, `arrows()`
 * This functions will add to a plot *without* refreshing or clearing the window

```{r}
plot(mat,  ylim = c(0,2), xlim = c(2014, 2021))
lines(x = mat + 1, type = "l")
```

```{r}
plot(mat, ylim = c(0,2), xlim = c(2014, 2019))
points(x = mat + 1, type = "p", col = 2)
```

```{r}
plot(mat, ylim = c(0,2), xlim = c(2014, 2019), type = "l")
abline(h = 0.4)
```

```{r}
plot(mat, ylim = c(0,2), xlim = c(2014, 2019), type = "l")
abline(v = 2016, lty = 2, col = "red")
```

```{r}
plot(x = c(0,1,2,3,4), y = c(0,2,4,5,10), type = "l")
abline(a = 0, b = 10/4, lty = 2, col = "red")
```

 * The `segments()` command takes a *from* coordinate (given as x0 and y0) and
a *to* coordinate (as x1 and y1) and draws the corresponding line.
 * `points()` takes two vectors of equal lengths with x and y values.
 * `arrow()` just like segments, where you provide a *from* coordinate (x0, y0) and a *to* coordinate (x1, y1)
 * As per the default behavior of `text()`, the string supplied as labels is *centered*
on the coordinates provided with the arguments x and y.

 
```{r}
plot(mat, ylim = c(0,2), xlim = c(2014, 2020), type = "l")
abline(h = c(0.1,1), col = "red", lty = 2,lwd = 2)
segments(x0 = c(2015, 2017),
         y0 = c(0.1, 0.1),
         x1 = c(2015, 2017),
         y1 = c(1, 1),
         col = "blue", lty = 3, lwd = 2)
arrows(x0 = 2015, y0 = 1.2, x1 = 2015.7, y1 = 0.3)
text(x = 2016.5, y = 1.3, labels = "Productivity Level: Low")
points(x = mat[,1][mat[,2] == 0.1],
       y = mat[,2][mat[,2] == 0.1], 
       pch = 19, cex =2, col = "green")
legend( "topright", #bottomleft, "topleft"
  legend = c("minmax", "2015-2017", "break"),
  pch = c(NA, NA, 19),
  lty = c(2, 3, NA), 
  col = c("red", "blue", "green"), 
  lwd = c(2, 2, NA))
```

## GGplot

 * Visualization with GGplot2
 
\tiny [Grammar of Graphics]{https://ggplot2.tidyverse.org/} 

* Base plotting is fast and simple for quick plots; ggplot2 is more powerful, consistent, and scalable for complex or professional visualizations.

# Ggplot - powerful alternatives to the standard plotting procedures in R written by Hadley Wickham.

 * Start by installing the ggplot2 package by calling `install.packages("ggplot2")`
 * Then load it `library(ggplot2)`
 * `ggplot2` functions like data in the *long* format (a column for every dimension, and a row for every observation)
 * Basic syntax of `qplot` is the same as earlier
 
 
```{r}
year <- c(2014, 2015, 2016, 2017, 2018, 2019, 2020)
energy_to_learn <- c(0.6, 0.5, 0.1, 0.7, 0.99, 0.5, 0.8) 
(mat <- cbind(year, energy_to_learn))

library(ggplot2)
qplot(x = mat[,1], y = mat[,2])
qplot(x = mat[,1], y = mat[,2], main = "Name", xlab = "X", ylab = "Y")
```


 * In `plot()`  we treated the graphics device as an *active canvas* where we added points, lines, and other features one by one.
 * *By contrast*, ggplot2 plots are stored as objects

```{r}
basic.plot <- plot(mat)
basic.plot # Since nothing is actually stored in the workspace

ggplot.plot <- qplot(x = mat[,1], y = mat[,2])
ggplot.plot
```

 * It opens up new ways to *modify or enhance* plots before displaying them


## How to build `ggplot` graphs:

 * `ggplot` graphics are built step by step by adding new elements
 * Elements of Grammar of graphics

 - **Data** - The data-set being plotted
 - **Aesthetics** - The scales onto which we plot our data (also, appearance and location)
 - **Geometry** - The visual elements used for our data
 - **Facet** - Groups by which we divide the data
 
```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()

```


```{r}
??geom
df.mat <- data.frame(V1 = mat[,1], V2 = mat[,2])

df.mat %>%
  ggplot() # to indicate that we want to create a plot.
```

### Building the *Axes*
 
- The next *layer* that we need to establish are the axes.
- In order to specify the axes, we need to use the `aes()` (aesthetic) function
- `aes()` is where we tell ggplot what columns we want to use

```{r}
df.mat %>%
  ggplot(aes(x = V1, y = V2))

# The same as
ggplot(data = df.mat, mapping = aes(x = V1, y = V2))
```

### Specifying *Geoms*

 - Setting Appearance Constants with `Geoms` - geometric modifiers:
 
 - `geom_line()` creates a line graph
 - `geom_point()` creates a scatter plot
 - `geom_bar()` creates a bar plot
 - etc.



 - We start to add *more layers* to it using a `+` sign


 * Simple scatterplot

```{r}
#??geom
df.mat %>%
ggplot(mapping = aes(x = V1, y = V2)) + 
  geom_point()

# Or just
ggplot(data = df.mat, aes(x = V1, y = V2)) + 
  geom_point()

# Connect by line
ggplot(data = df.mat,mapping = aes(x = V1, y = V2)) + 
  geom_point() +
  geom_line(data = df.mat %>% filter(V1 < 2019), mapping = aes(x = V1, y = V2))
```

 * Be careful with the position of `+` sign
 
```{r}
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
  geom_point() 
  + geom_line()
```

 * Anything you put in the `ggplot()` function can be seen by any `geom` layers that you add (i.e., these are *universal plot* settings).
 
 * You can also specify aesthetics for a given `geom` *independently* of the aesthetics defined globally in the `ggplot()` function.
 
```{r}
df.mat %>% 
  ggplot() + 
    geom_point(aes(x = V1, y = V2))+ 
    geom_line(aes(x = V1, y = V2))
```
 


### Adding a Title and Axis Labels

 - Graph keeps the column names as the labels for both of the axes.
 - To change the *axis labels* for a plot, we can use the `labs()` function and add it as a *layer* onto the plot
 - `labs()` can change *both* the *axis* labels as well as the *title* 

```{r}
df.mat %>%
ggplot(aes(x = V1, y = V2)) + 
  geom_point() +
  geom_line() +
  labs( title = "Energy to learn from 2014 to 2020", 
        y = "Energy to learn", 
        x = "Year")
```

 * pch - shape
 * cex - size

 * No need to adjust limits:

```{r}
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
  geom_point() +
  geom_line() +
  geom_point(aes(x = 2022, y = 2), size = 5, shape = 7)
```


```{r}
# compatible
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
  geom_point(cex = 3, pch = 6, color = "blue") + 
  geom_line(color = "red", lty = 2)

ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
  geom_point(alpha = 0.5, size = 3, color = "coral")
```

 * No longer requires to rerun every plotting command - just store and reuse

```{r}
(plot.one <- ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_line(color = "darkgreen", linetype = 2))

plot.one + geom_point(size = 3, shape = 6, color = "coral") 
plot.one + geom_point(size = 4, shape = 1, color = "coral") 
```

 
 * If you split a data set into categories using a *factor* object, `ggplot2` can *automatically* apply particular styles to different categories. 

```{r}
df.mat
df.mat$V3 <- c("H", "L", "M")[sample(x = 3, size = 7, replace = T)]
df.mat$V3 <- factor(df.mat$V3)

plot(x = df.mat$V1, y = df.mat$V2, pch = df.mat$V3)

ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point(aes(shape = V3), size = 2)
```

```{r}
plot(mat, ylim = c(0,2), xlim = c(2014, 2020), type = "l")
abline(h = c(0.1,1), col = "red", lty = 2,lwd = 2)
segments(x0 = c(2015, 2017),
         y0 = c(0.1, 0.1),
         x1 = c(2015, 2017),
         y1 = c(1, 1),
         col = "blue", lty = 3, lwd = 2)
arrows(x0 = 2015, y0 = 1.2, x1 = 2015.7, y1 = 0.3)
text(x = 2016.5, y = 1.3, labels = "Productivity Level: Low")
points(x = mat[,1][mat[,2] == 0.1],
       y = mat[,2][mat[,2] == 0.1], 
       pch = 19, cex =2, col = "green")
legend( "topright", #bottomleft, "topleft"
  legend = c("minmax", "2015-2017", "break"),
  pch = c(NA, NA, 19),
  lty = c(2, 3, NA), 
  col = c("red", "blue", "green"), 
  lwd = c(2, 2, NA))
```

```{r}
df.mat$V3 <- factor(c("good", "good", "break", "good", "good", "good", "good"))
# The same plot
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point(aes(shape = V3, color = V3), size = 2)+
  geom_line(color = "black", lty = 2) + 
  geom_hline(yintercept = c(0.1,1), color="red") + 
  geom_segment(x = 2015, y = 0.1, xend = 2015, yend = 1, color = "red", linetype = 3) +
  geom_segment(x = 2017, y = 0.1, xend = 2017, yend = 1, color = "red", linetype = 3)
```

 * Zoom in
 
```{r}
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
  geom_point()+
  scale_x_continuous(limits = c(2015, 2019)) +
  scale_y_continuous(limits = c(0, 0.5)) 
```
 

 * Categorical for color should be factor variable
  - `factor(gear)` will determine the size of the dots.  
  -  `factor(carb)` will determine the color of the dots

```{r}
mtcars %>%
  ggplot(aes(x = mpg, y = disp, col = gear)) +
  geom_point()

summary(mtcars)

mtcars %>%
  ggplot(aes(x = mpg, y = disp, col = factor(gear))) +
  geom_point()

mtcars %>%
  ggplot(aes(x = mpg, y = disp, col = factor(carb), size = factor(gear))) +
  geom_point()
```

 * Using `Facets` in ggplot2
 
 - Facet is a way in which you can add additional categorical variables to your plot.  
 - The facet helps in building the chart by dividing the data into two or more *groups*.
 - Use tilder(`~`)

```{r}
ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point( size = 2)+
  geom_line(color = "black", lty = 2) + 
  facet_grid(.~V3)


ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point(aes(shape = V3, color = V3), size = 2)+
  geom_line(color = "black", lty = 2) + 
  facet_grid(V3~.)
```



```{r}
plot.two <- ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point(aes(shape = V3, color = V3), size = 2) +
            geom_line(color = "black", lty = 2) + 
            theme(legend.position = "None")


plot.three <- plot.two + 
              labs(caption = paste( "* Note 1: Add text ",
                       "\n** Note 2: Text here.",
                       "\n*** Note: ", paste(nrow(df.mat), "rows")))

plot.three

plot.three + theme(plot.caption = element_text(hjust = 0,  face = "italic", size = 6.5))
```


### Other plots on real data

#### Bar plot
 
```{r}
load("Lesson 5.rda")
covid.subset

(g1 <- covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval)) +
      geom_bar())

cat(str_wrap("Some long string", width = 0.5))
cat(str_wrap("Some long string", width = 1))
cat(str_wrap("Some long string", width = 9))

(g2 <- g1 +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)))
```

 * Use the `fill` attribute in the `geom_bar()` function to give the bar plot a color.
 
```{r}
# Bar colors
covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval)) +
      geom_bar(fill = "darkgreen") +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15))

covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q8.1.SourceEval)) +
      geom_bar() +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15))


covid.subset$Q8.1.SourceEval <- DescTools::reorder.factor(covid.subset$Q8.1.SourceEval, new.order = c(2:4,1))

covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q8.1.SourceEval)) +
      geom_bar() + 
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme(legend.position = "None")
```


 * Pallet
 
```{r}
covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q8.1.SourceEval)) +
      geom_bar() +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
      scale_fill_brewer(palette = "Accent")
```

[RBrewColor Pallet chart](Lesson 5.png)


```{r}
library(RColorBrewer)
head(brewer.pal.info[1:2], 20)
```

See more in `wesanderson` package. 

 * The attribute `position` is given as `fill`, i.e. we’ll get the bar plot in terms of proportion.

```{r}
covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q2.Gender)) +
      geom_bar() +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15))

# showing proportion instead of count
(g3 <- covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q2.Gender)) +
      geom_bar(position = "fill") +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15)))

g3 + labs(fill = "Gender", x = "Source Evaluation", y = "")

# Dodge comparison
(g4 <- covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval, fill = Q3.Age_group)) +
      geom_bar(position = "dodge") +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15)))
```


 *  Combine graphs with `grid.arrange` from `gridExtra` 
 
```{r fig.width=10}
# First task
gridExtra::grid.arrange(g3, g4, ncol = 2)
``` 

 * `coord_cartesian()` - default coordinate system in ggplot2
 * `coord_flip()` - for horizontal graphs

```{r fig.width=8}
g4 + facet_grid(.~Q2.Gender)
g4 + facet_grid(.~Q2.Gender) + coord_cartesian()

g4 + facet_grid(Q2.Gender~.)
g4 + facet_grid(Q2.Gender~.) + coord_flip()
g4 + facet_grid(~Q2.Gender) + coord_flip()
```

```{r}
g4 + facet_grid(Q17.Mental~Q2.Gender, scales = "free_y") 
```


* Support for *statistical transformation* in ggplot and *themes*
 
```{r}
covid.subset  %>% 
      ggplot(mapping = aes(x = Q8.1.SourceEval)) +
      stat_count(fill = "darkgreen") +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15))

covid.subset  %>% 
      ggplot() +
      geom_bar(aes(x = Q8.1.SourceEval, y = ..prop.., group = 1), fill = "darkgreen") +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 15))

ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point()+
            geom_line(color = "black", lty = 2) + 
            stat_smooth(method = "lm", se = F)

ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point()+
            geom_line(color = "black", lty = 2) +
            geom_hline(yintercept = mean(df.mat$V2))
```

```{r}
library(gridExtra)
tm <- ggplot(data = df.mat, mapping = aes(x = V1, y = V2)) + 
            geom_point()

grid.arrange(tm + theme_bw(), 
             tm + theme_linedraw(),
             tm + theme_dark(),
             tm + theme_minimal(),
             tm + theme_classic(),
             tm + theme_gray(),
             ncol = 3, nrow = 2)
```
 
#### Boxplot

 - A box plot can be created by using `geom_boxplot()`
 
```{r}
airquality  

(g5 <- airquality  %>% 
      ggplot(mapping = aes(x = Month, y = Temp)) +
      geom_boxplot())

(g5 <- airquality  %>% 
      ggplot(mapping = aes(x = factor(Month), y = Temp)) +
      geom_boxplot())

tapply(airquality$Temp, airquality$Month, mean)

airquality$HalfMonth <-  ifelse(airquality$Day > 15, "first", "second")
# Second task
(g5.1 <- airquality  %>% 
      ggplot(mapping = aes(x = factor(Month), y = Temp, fill = factor(HalfMonth))) +
      geom_boxplot())

g5 +
  geom_jitter() # have a better idea of the number of points


g5 +
  geom_jitter(alpha = 0.2, color = "tomato") 

# interactive plots
plotly::ggplotly(g5)
```

#### Line Plot

 * Make X-axis as Date
 
```{r}
theme_set(theme_classic())
library(readxl)
UR.gender <- read_excel("Lesson 5.xlsx", sheet = 2)

UR.gender %>% 
  filter(Metric != "Total") %>%
  mutate(Year = as.Date(paste(Year, "01-01", sep = "-"))) %>%
  ggplot(aes(x = Year, y = UR , col = Metric)) + 
  geom_line(size = 1)+
  theme(legend.position = "bottom") + labs(col = "")+
  ggtitle("Fig ??: Unemployment rate by gender")
```

```{r}
UR.marz <- read_excel("Lesson 5.xlsx", sheet = 3)

UR.marz <- UR.marz %>% tidyr::gather(Year, Unemployment,  -Region) %>% arrange(Region, Year)
UR.marz$Year <- as.Date(paste(UR.marz$Year, "01-01", sep = "-"))


UR.marz %>% ggplot(aes(x = Year, y = Unemployment , col = Region)) + 
  geom_line(size = 1)


UR.marz.lab <- UR.marz[UR.marz$Year=="2018-01-01", ]

UR.marz %>% ggplot(aes(x = Year, y = Unemployment , col = Region)) + 
  geom_line(size = 1) +
  geom_text(data = UR.marz.lab, mapping = aes(label = Region), color = "black", size = 3) + 
  ggtitle("Fig #??: Unemployment rate by marzes (2012-2019)")+
  ylab("%") + xlab("Year") + theme(legend.position = "None")

# Task 4
# Or Faceting
UR.marz %>% ggplot(aes(x = Year, y = Unemployment , col = Region)) + 
  geom_line(size = 1) + theme(legend.position = "None")+
  facet_wrap(~Region)
```


#### Correlogram and save plots

```{r}
theme_set(theme_gray())
# Correlation matrix
corr <- round(cor(mtcars), 1)

# Plot
library(ggcorrplot)
ggcorrplot(corr, title = "Correlogram of mtcars")

g1 <- ggcorrplot(corr, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method = "circle", 
           colors = c("tomato2", "white", "springgreen3"), 
           title = "Correlogram of mtcars", 
           ggtheme=theme_bw)

getwd()
ggsave(filename = "Lesson-5-Plot-1.png", plot = g1)
```
